Light inextensible strings (thread) under tension in the Schwarzschild geometry

### 2021-02-26 tpe_edreview_rebuttal.txt

On reflection the dominant energy condition is not really an
appropriate or informative or useful or interesting concept for the
intended readership.  The important thing is that tension greater $\mu
c^2$ is unphysical and this context, the unphysicality is manifested
by transverse waves travelling faster than light.

In answer to the question, we may take the stress-energy tensor for a
stretched string to be diagonal with elements $\mu,-T,0,0$.  Then the
dominant energy condition is that, for any future-pointing causal
vector, $-T^\mu_\nu V^\nu$ must also be a future-pointing causal
vector.  We can see directly that if $T>\mu$ the DEC is violated.

But this is a massive digression to the main thrust of the manuscript
and any adequate pedagogical explanation would be an intrusive
distraction.  It is like using a slegehammer to crack a nut.  I have
removed any reference to the DEC on these grounds, and further
emphasised the unphysicality of the wave speed being $>c$.

> 
> 5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
> circular string at constant r will, trivially, have proper length
> $2\pi r$."  Stating the relation this way muddies the logic and the
> distinction between flat and Schwarzschild space.  The r-coordinate of
> a spherical shell centered on a point mass M is defined operationally
> by measuring the circumference of the shell and dividing the
> circumference's value by $2\pi$.  That must be understood before the
> calculations of Eq. (2) can make sense.  As the sentence is presently
> structured it may obscure the distinction between the r coordinate and
> radial lengths because it invites the undergraduate reader to
> interpret the "triviality" as a statement about plane Euclidean
> geometry instead of the Schwarzschild geometry's definition of the
> r-coordinate.
>

The referee is absolutely correct here and raises an important point.
When I said that the circumference was "trivially" $2\pi r$, what I
meant was that integrating equation (1) from $phi=0$ to $2\pi$ at
fixed $t,r,\theta$ was trivial.  This is a direct consequence of the
*definition* of "r" [which is sometimes called the circumferential
radial coordinate].  I remember struggling with this myself as an
undergraduate (I wanted to know what the "real" radius was ...).
Einstein himself said "it is not so easy to free oneself from the idea
that coordinates must have an immediate metrical meaning", so I have
good company.

On reflection this point in the manuscript represents a good
"teachable moment".

> 6.  Question.  There may be a typographical error in Eq. (2).
> Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
> written the equation is not dimensionally consistent, since r and M
> are measured in the same units.  Compare to Taylor & Wheeler's
> Exploring Black Holes, p. 2-28, Eq. (27).
>

### 2021-02-26 tpe_edreview_rebuttal.txt

On reflection the dominant energy condition is not really an
appropriate or informative or useful or interesting concept for the
intended readership.  The important thing is that tension greater $\mu
c^2$ is unphysical and this context, the unphysicality is manifested
by transverse waves travelling faster than light.

In answer to the question, we may take the stress-energy tensor for a
stretched string to be diagonal with elements $\mu,-T,0,0$.  Then the
dominant energy condition is that, for any future-pointing causal
vector, $-T^\mu_\nu V^\nu$ must also be a future-pointing causal
vector.  We can see directly that if $T>\mu$ the DEC is violated.

But this is a massive digression to the main thrust of the manuscript
and any adequate pedagogical explanation would be an intrusive
distraction.  It is like using a slegehammer to crack a nut.  I have
removed any reference to the DEC on these grounds, and further
emphasised the unphysicality of the wave speed being $>c$.

> 
> 5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
> circular string at constant r will, trivially, have proper length
> $2\pi r$."  Stating the relation this way muddies the logic and the
> distinction between flat and Schwarzschild space.  The r-coordinate of
> a spherical shell centered on a point mass M is defined operationally
> by measuring the circumference of the shell and dividing the
> circumference's value by $2\pi$.  That must be understood before the
> calculations of Eq. (2) can make sense.  As the sentence is presently
> structured it may obscure the distinction between the r coordinate and
> radial lengths because it invites the undergraduate reader to
> interpret the "triviality" as a statement about plane Euclidean
> geometry instead of the Schwarzschild geometry's definition of the
> r-coordinate.
>

The referee is absolutely correct here and raises an important point.
When I said that the circumference was "trivially" $2\pi r$, what I
meant was that integrating equation (1) from $phi=0$ to $2\pi$ at
fixed $t,r,\theta$ was trivial.  The resulting proper string length is
a direct consequence of the *definition* of "r" [which is sometimes
called the circumferential radial coordinate].

This comment has made me realise that the $2\pi r$ proper string
length is actually the simplest proper length calculation possible in
the Schwarzschild metric, and I have rephrased accordingly, and I hope
it is more instructive.  I remember struggling with this myself as an
undergraduate (I wanted to know what the "real" radius was ...).
Einstein himself said "it is not so easy to free oneself from the idea
that coordinates must have an immediate metrical meaning", so I have
good company!

> 6.  Question.  There may be a typographical error in Eq. (2).
> Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
> written the equation is not dimensionally consistent, since r and M
> are measured in the same units.  Compare to Taylor & Wheeler's
> Exploring Black Holes, p. 2-28, Eq. (27).
>

### 2021-02-26 string_physics_educator_revision3.tex
fixes #38

\noindent Here I consider the spatial component~$ds^2=
dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
circular string at constant~$r$ will, trivially, have proper
length~$2\pi r$.  Considering first the radial component we may follow
Taylor and Wheeler \cite{taylor2000} and calculate the proper length
$\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant
2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:

>>>>

\noindent Here I consider the spatial component~$ds^2=
dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.

The simplest configuration for a string would be circular at
constant~$r$ whose proper length would be given by integration of
equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
(this is why $r$ is sometimes called ``circumferential radius'').
Note carefully that $r$ is not the distance {\em to} anything, so this
result---which operates in Schwarzschild geometry and not plane
Euclidean geometry---is less obvious than it might appear.
Considering next a purely radial string we may follow Taylor and
Wheeler \cite{taylor2000} and calculate the proper length
$\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant
2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:
